Vibronic Coupling Effects in the Photoelectron Spectrum of Ozone: A Coupled-Cluster Approach

### 2024-09-04 article.tex
Submitted: Dr. Krylov's edits

\documentclass[
12pt,
a4paper,
prb,
superscriptaddress,
% tightenlines,  % single-spaced
]{revtex4}
\usepackage{amsmath,amssymb,physics}
\usepackage{graphicx} % For figures with images
\usepackage{natbib} % bibliography
% \setcitestyle{super, sort&compress, comma} % chemistry style (prb's default)
\usepackage{hyperref} % Clickable links in the document and their colors
\hypersetup{
    colorlinks=true, % false: boxed links; true: colored links
    linkcolor=cyan, % color of internal links
    citecolor=magenta, % color of links to bibliography
    filecolor=magenta, % color of file links
    urlcolor=cyan, % color of external links
    runcolor=cyan
}

@@

\author{Pawe{\l} W{\'o}jcik}
\affiliation{Department of Chemistry, University of Southern California, Los Angeles, California 90089, USA}
\affiliation{Laboratory of Theoretical Chemistry, Institute of Chemistry, ELTE
E{\"o}tv{\"o}s Lor{\'a}nd University, P{\'a}zm{\'a}ny P{\'e}ter stny. 1/A,
Budapest, Hungary}

\author{Anna I. Krylov}
\affiliation{Department of Chemistry, University of Southern California, Los Angeles, California 90089, USA}

\author{Hannah Reisler}
\affiliation{Department of Chemistry, University of Southern California, Los Angeles, California 90089, USA}

\author{P{\'e}ter G. Szalay}
\affiliation{Laboratory of Theoretical Chemistry, Institute of Chemistry, ELTE
E{\"o}tv{\"o}s Lor{\'a}nd University, P{\'a}zm{\'a}ny P{\'e}ter stny. 1/A,
Budapest, Hungary}

\author{John F. Stanton}
\email[Electronic address: ]{jfstanton137@gmail.com}
\affiliation{Quantum Theory Project, Departments of Chemistry and Physics, University of Florida, Gainesville, FL, USA 32611}
\affiliation{Laboratory of Theoretical Chemistry, Institute of Chemistry, ELTE
E{\"o}tv{\"o}s Lor{\'a}nd University, P{\'a}zm{\'a}ny P{\'e}ter stny. 1/A,
Budapest, Hungary}

@@

\begin{abstract}
One of the most important areas of application for equation-of-motion coupled
cluster (EOM-CC) theory is the prediction, simulation and analysis of
various types of electronic spectra.   In this work, the EOM-CC method known
as EOMIP-CC is applied to the closely lying and coupled pair of states of the
ozone cation -- ${\tilde X}^2A_1$ and ${\tilde A}^2B_2$ -- using sophisticated
approaches that extend to the full singles, doubles, triples and quadruples
model (EOMIP-CCSDTQ).   Combined with a venerable and powerful method for
calculating vibronic spectra from the Hamiltonian produced by EOMIP-CC
calculations, this research provides a spectrum that is in good agreement with
the photoelectron spectrum of ozone.   An important result here is that the
calculations suggest that the adiabatic gap separating these two electronic
states is somewhat smaller than currently thought; an assignment of the
simulated spectrum together with the more precise band positions of the
experimental measurements suggests that this energy gap is 1366$\pm$65
cm$^{-1}$.
\end{abstract}

@@

Amongst the brotherhood of triatomic molecules, it cannot be denied that water
(H$_2$O) is the most important, the most highly studied, and the most
well-understood.  Beyond H$_2$O, there are many triatomic molecules that have
an environmental, technological or biological importance, have been subjected
to many studied and are understood to various levels of detail.  Perhaps the
most interesting such case is ozone (O$_3$), which has a vast number of
important properties, a very rich history of study \cite{chappuis}, and 
unlike the relatively simple water molecule, a profound quantum-mechanical
complexity \cite{Babikov:anomalousOzone:2003}.  In the latter context, while
we think of (and an NMR experiment would reveal) as two distinct kinds of
oxygen atoms in ozone, the full molecular Hamiltonian does not distinguish
between them types, with the three equivalent structures separated by a
barrier that lies tantalizingly close to the O$_3$ $\rightarrow$ O$_2$ + O
dissociation threshold (102.4 kJ/mol) \cite{Ruscis:ATcT:2022}. In reality, the
energy levels of ozone all have a near triple ($e+a$) degeneracy, albeit with
a tunneling splitting so small that it can be ignored, along with a
semi-infinite lifetime (despite opposing opinion
\cite{Boggs:BerryOzone:2006}).   More than a half century ago, this intriguing
aspect of the ozone molecule was first discussed by Berry
\cite{Berry:Ozone:1960}.

Beyond the structural aspects of ozone, other mysteries center around this
curious molecule.  For example, the distribution of the eighteen distinct
$^{16}$O/$^{17}$O/$^{18}$O isotopologues in the Earth's atmosphere differs
from what is expected based on the natural isotopic abundance, a puzzle that
has been open for more than three decades
\cite{Mauersberger:OzoneMystery:1990}.

Among quantum chemists, ozone has a notorious history, with its strong
biradical character causing significant difficulties in calculations of its
ostensibly simple ground state molecular properties.  An early 1989 study
\cite{Stanton:Ozone:1989, Stanton:Ozone:1989b} by the Bartlett group and
collaborators found that the CCSD+T(CCSD) method predicted that the molecular
equilibrium structure of ozone would have $C_s$ symmetry (that is, the
asymmetric stretching harmonic frequency predicted by this method is
imaginary), a finding that led to a search for better treatment of
non-iterative triple excitations, ultimately leading to the well-known CCSD(T)
treatment \cite{Raghavachari:89, Urban:ccsd(t):1985, GenrefCCSD(T):93}. While
CCSD(T) and higher-level coupled-cluster methods available today
\cite{kucharski:ccsdtq:1992, Kallay:CCHigh, Matthews:ncc:2015} do a good job
in predicting the molecular vibrational potential, an elaborate multireference
configuration interaction study by Dawes {\it et al.} has done an excellent
job on the entire ground-state surface out to the dissociation limit
\cite{Dawes:ozone:2013}. As such, the quantum chemical understanding and
fidelity for the ground electronic state of O$_3$ is now at a mature level.

Qualitatively, the challenge posed to electronic structure theory by ozone
ultimately arises from its closely spaced highest-occupied ($a_2$) and
lowest-unoccupied ($b_1$) molecular orbitals (HOMO and LUMO, respectively).
The two electron configurations $[\cdots]b_2^2 a_1^0$ and $[\cdots]b_2^0
a_1^2$ mix strongly, posing the aforementioned challenges with (especially
single-reference) quantum chemical methods.   A second feature consequence of
the identity and energetic proximity of these two orbitals is that the ozone
cation (which is isoelectronic to the NO$_2$ radical) has closely lying
$^2A_1$ and $^2B_2$ electronic states.   Like the associated states in NO$_2$,
both of these states are plagued by orbital symmetry-breaking problems
\cite{Davidson:SymmBreak:76}, a problem that greatly complicates
quantum-chemical efforts to characterize them.   One of the many
accomplishments of the Bartlett group has been the integral role played by
them in the development of equation-of-motion coupled-cluster theory
\cite{Stanton:93:EOMCC, Nooijen:EOMEA:95, Bartlett:CC_review:07} (EOM-CC, also
known as linear response coupled-cluster theory \cite{Koch:90:LinResp}). These
methods provide a very efficient and simple way to study certain classes of
what are often termed ``multireference problems" \cite{Krylov:EOMRev:07}, and
are ideally suited to studying many reactive intermediates, radicals,
biradicals and electronically excited states.   From a somewhat wider
viewpoint, the existence of closely spaced electronic states always carries
the potential for (possibly strong) vibronic coupling, a phenomenon that can
play an important role in molecular dynamics and spectroscopy.  Indeed, one of
the great successes of EOM-CC methods has been in their ability -- if and only
if combined with vibronic coupling models -- to enable high-quality
simulations of complicated electronic spectra.   Such work provides important
insight into the nature of vibronic coupling in molecular systems, as has been
exemplified by various application studies (for example, see
Refs.~\cite{Stanton:NO3:07} and~\cite{Koppel:02}).

@@

\begin{figure}
    \includegraphics[width = 16 cm]{./figures/ozone_intro}
    \caption{ 
        Two lowest states of the ozone cation and normal modes of the ozone
        neutral.
    }
    \label{fig:ozone_intro}
\end{figure}

\section{Theory}

Ozone is a $C_{2v}$ molecule (following Mulliken's
convention~\cite{Mulliken:55:symnot} the molecule is placed in the $yz$-plane
and the molecular symmetry axis aligns with the $z$ direction) with three
normal modes: symmetric stretch, symmetric bend, and asymmetric stretch.  The
asymmetric stretch is of $b_2$ symmetry. The two lowest electronic states of
the ozone cation are close in energy. The lower state is fully symmetric while
the higher one is of $B_2$ symmetry, the same symmetry as the asymmetric
stretch. The close energetic separation and the matching symmetry of the
asymmetric stretch is accompanied by significant vibronic effects in the
lowest two states of the ozone photoelectron.  Figure~\ref{fig:ozone_intro}
presents a graphical summary of this information.

The vibronic states of the ozone cation are simulated using the model
Hamiltonian of K{\"o}ppel, Domcke, and Cederbaum (KDC
Hamiltonian).\cite{Cederbaum:LVC:84,KDC:81,Koppel:CIbookCh7:04} This is a
multi-state and multi-mode Hamiltonian defined in the basis of diabatic
states. The ozone cation is considered in the basis of two quasi-diabatic
states coupled by one mode (mode number 3). The model also includes two
symmetric modes (modes number 1 and 2)
\begin{subequations}
    \begin{equation}
        H = H _0 \mathbf{1}
        +
        \begin{pmatrix}
            V ^{(1)}  & \lambda _3 Q _3\\
            \lambda _3 Q _3 & V ^{(2)}
        \end{pmatrix}
    \end{equation}
    \begin{equation}
        H _0 = 
        \frac{1}{2} \left(\sum _{i = 1}^3 
        - \omega _i \frac{\partial ^2}{\partial Q _i ^2 }\right)
        + \frac{1}{2}\omega _3 Q _3 ^2
    \end{equation}
    \begin{equation}
        V ^{(\alpha)} = 
        E ^{(\alpha)} 
        + \sum _{i \in \{1,2\}} 
            \kappa ^{(\alpha)} _i Q _i 
        + \sum _{i,j \in \{1,2\}} 
            \kappa ^{(\alpha)} _{ij} Q _i Q _j 
        + \sum _{i,j,k \in \{1,2\}} 
            \kappa ^{(\alpha)} _{ijk} Q _i Q _j Q _k 
        + \sum _{i,j,k,l \in \{1,2\}} 
            \kappa ^{(\alpha)} _{ijkl} Q_i Q _j Q _k Q _l.
    \end{equation}
    \label{eq:KDC}
\end{subequations}
The vibronic Hamiltonian elements are expanded around the equilibrium geometry
of neutral ozone. $E ^{(\alpha)}$ are the vertical ionization energies
calculated at that geometry. $Q_i$ are the dimensionless normal coordinates of
ozone, $\kappa$ are the coefficients of expansion of the potential along the
fully symmetric coordinates, and $\lambda$ the linear diabatic coupling,
$\omega$ represent the harmonic frequencies of the neutral.

The parameters that enter the KDC Hamiltonian are computed using \emph{ab
inito} coupled-cluster (CC) methods and its equation-of-motion EOM-CC
extension.~\cite{Bartlett:CC_review:07, Krylov:EOMRev:07, Bartlett:Book:09,
Christiansen:EOMRev:11, Bartlet:EOMRev:12, Krylov:OSRev} The CC expansion is
truncated to singles and doubles (CCSD), singles, doubles and triples (CCSDT)
as well as singles, doubles, triples and quadruples
(CCSDTQ).\cite{Matthews:ncc:2015}  The states of the cation are described with
EOM-CC for ionization potential (EOMIP).~\cite{StantonGauss:EOMIP:99} The
definition of quasi-diabatic states of Ichino, Gauss and Stanton based on the
EOMIP-CC method (EOMIP-CC-QD) is used here~\cite{Stanton:EOMIPdeg:09}. All CC
and EOM-CC calculations correlate all electrons, in other words, the
frozen-core approximation is not invoked.

The geometry of the neutral ozone and the computation of its harmonic
frequencies and normal coordinates uses
CCSDT/ANO1.~\cite{Almlof:ANO,Almlof:ANO:1988} 
The linear diabatic coupling $\lambda$ is computed at the same geometry using
EOMIP-CCSD-QD/ANO1. The expansion coefficients $\kappa$ are computed using the
EOMIP-CCSD/ANO1 energies calculated on a grid around the same geometry.

The vertical ionization energies are calculated using a composite method.  The
starting value is the complete basis set (CBS) extrapolation of the
EOM-IP-CCSDT/cc-pCVnZ energies with n = 5, 6.~\cite{Woon:95:CCBS} These values
are augmented with two corrections: the $\Delta$Q correction in the cc-pwCVTZ
basis set and the relativistic correction calculated using
EOM-IP-CCSD/cc-pwCVTZ.~\cite{Dunning:02:p(w)CVnZ} The error estimate of the
extrapolated CBS reference values is reported using half of the absolute value
of the difference between the best \emph{ab initio} value and the extrapolated
value.  For the remaining corrections (quadruples correction, $\Delta Q$, and
the relativistic correction) the error estimate is reported as half of the
absolute value of the correction.

To compare the simulated photoelectron spectrum to experiment, the
cross-section ratio A$_1$:B$_2$ of $1$:$1.35$ is used.\cite{KDC:O3:92}
Additionally, the stick spectrum is broadened with the Lorenzian envelopes
normalized to the feature intensities
\begin{equation}
    f _{envelope}(x, x _i, I _i) = 
    \frac{I_i}{(x-x _i)^2 + (\gamma/ 2) ^2}.
    \label{eq:lorentzian}
\end{equation}
$x _i$ is the position of the spectral peak, $I _i$ is its intensity and
$\gamma$ is the peak's width.

The spectrum is simulated using the \textsc{xsim} program in \textsc{CFOUR},
with the basis of 50 harmonic states in each mode and 6000 iterations of the
Lanczos procedure.~\cite{Sharma:xsim_socjt:2024} All calculations are
completed using \textsc{CFOUR}.~\cite{cfour, cfour:2020}

\section{Results}

The optimized geometry of ozone gives a bond length of $1.270$~\AA{} and a
bond angle of $116.9^\circ$. The two symmetric normal modes have frequencies
$1169$~cm$^{-1}$ and $724$~cm$^{-1}$, while the asymmetric stretch has a
frequency equal to $1129$~cm$^{-1}$. The computed value of the linear diabatic
coupling constant $\lambda$ is $1394$~cm$^{-1}$. The vertical ionization
energy for the first excited state, $E^{(1)}$, computed using the composite
method described above is  $12.827$~eV. Our error estimate for this value is
$30$~meV, see Table~\ref{tab:vertical_ionization_energy} for details.  We note
that the convergence of the vertical gap between the two states is much faster
and this value is estimated as $123\pm8$~meV, see Table~\ref{tab:vertical_gap}
for details.

\begin{table}
    \caption{
        Vertical ionization energies with the error estimates, eV.
    }
    \label{tab:vertical_ionization_energy}
    \begin{center}
        \begin{tabular}[c]{|l|rr|r|}
            \hline
            Contribution  & $^2$A$_1$ & $^2$B$_2$ & Error estimate \\ \hline
            CBS           & 12.872    & 12.981    & 0.02 \\
            $\Delta$Q     & -0.037    & -0.021    & 0.02 \\
            Relativistic  & -0.008    & -0.010    & 0.005 \\ \hline
            Sum           & 12.827    & 12.950    & 0.03 \\ \hline
        \end{tabular}
    \end{center}
\end{table}

\begin{table}
    \caption{
        The energy gap between vertical ionization energies of the $^2$A$_1$
        and $^2$B$_2$ states with error estimates, meV. See caption of
        Table~\ref{tab:vertical_ionization_energy}.
    }
    \label{tab:vertical_gap}
    \begin{center}
        \begin{tabular}[c]{|l|r|r|}
            \hline
            Contribution             &  Vertical gap    & Error estimate \\ \hline
            EOM-CCSDT/CBS            &  108.8           & 0.9 \\
            $\Delta$Q/pwCVTZ         &   16.5           & 8 \\
            Relativistic/CCSD/pwCVTZ &   -2.2           & 1.1 \\ \hline
            Final value, meV         &  123             & 8 \\
            Final value, cm$^{-1}$   &  990             & 65 \\ \hline
        \end{tabular}
    \end{center}
\end{table}

>>>>

%\documentclass[12pt,a4paper,prb,superscriptaddress,
% tightenlines,  % single-spaced
%]{revtex4}
\documentclass[12pt,prb,aps]{revtex4}
\usepackage{amsmath,amssymb,physics}
\usepackage{graphicx} % For figures with images
\usepackage{natbib} % bibliography
% \setcitestyle{super, sort&compress, comma} % chemistry style (prb's default)
\usepackage{hyperref} % Clickable links in the document and their colors
\hypersetup{
    colorlinks=true, % false: boxed links; true: colored links
    linkcolor=cyan, % color of internal links
    citecolor=magenta, % color of links to bibliography
    filecolor=magenta, % color of file links
    urlcolor=cyan, % color of external links
    runcolor=cyan
}

@@

\author{Pawe{\l} W{\'o}jcik$^a$,  Hanna Reisler$^a$, P{\'e}ter G. Szalay$^b$, Anna I. Krylov$^{a,\dagger}$, John F. Stanton$^{b,c,*}$\\
{\small $^a$ Department of Chemistry, University of Southern California, Los Angeles, California 90089, USA}\\
{\small $^b$  Laboratory of Theoretical Chemistry, Institute of Chemistry, ELTE
E{\"o}tv{\"o}s Lor{\'a}nd University, P{\'a}zm{\'a}ny P{\'e}ter stny. 1/A,
Budapest, Hungary}\\
{\small $^c$ Quantum Theory Project, Departments of Chemistry and Physics, University of Florida, Gainesville, FL, USA 32611}\\
{$\dagger$ krylov@usc.edu; $^*$ jfstanton137@gmail.com}}

@@

\begin{abstract}
One of the most important areas of application for equation-of-motion coupled
cluster (EOM-CC) theory is the prediction, simulation, and analysis of
various types of electronic spectra. 
In this work, the EOM-CC method for ionized states, known
as EOM-IP-CC, is applied to the closely lying and coupled pair of states of the
ozone cation --- ${\tilde X}^2$A$_1$ and ${\tilde A}^2$B$_2$ --- using highly accurate 
treatments including up to the full single, double, triple, and quadruple excitations
(EOM-IP-CCSDTQ). Combined with a venerable and powerful method for
calculating vibronic spectra from the Hamiltonian produced by EOM-IP-CC
calculations, the simulations yield a spectrum that is in good agreement with
the photoelectron spectrum of ozone. Importantly,  the
calculations suggest that the adiabatic gap separating these two electronic
states is somewhat smaller than currently thought; an assignment of the
simulated spectrum together with the more precise band positions of the
experimental measurements suggests that this energy gap is 1,366$\pm$65
cm$^{-1}$.
\end{abstract}

@@

Amongst the brotherhood of triatomic molecules, it cannot be denied that water
(H$_2$O) is the most important, the most highly studied, and the most
well-understood. Beyond H$_2$O, many triatomic molecules that have
an environmental, technological or biological importance have been subjected
to many studies and are understood to various levels of detail. Perhaps the
most interesting such case is ozone (O$_3$), which has a vast number of
important properties, a very rich history of study \cite{chappuis}, and 
unlike the relatively simple water molecule, a profound quantum-mechanical
complexity \cite{Babikov:anomalousOzone:2003}. Structurally, while
we think of ozone as having two distinct kinds of oxygen atoms (and an NMR experiment would reveal that), the full molecular Hamiltonian does not distinguish
between them. Rather, it yields the three equivalent structures separated by a
barrier that lies tantalizingly close to the O$_3$ $\rightarrow$ O$_2$ + O
dissociation threshold (102.4 kJ/mol) \cite{Ruscis:ATcT:2022}, as shown in 
Fig. \ref{fig:3fold}.
In reality, the
energy levels of ozone all have a near triple ($e+a$) degeneracy, albeit with
a tunneling splitting so small that it can be ignored, along with a
semi-infinite lifetime (despite opposing opinion
\cite{Boggs:BerryOzone:2006}). More than a half century ago, this intriguing
aspect of the ozone molecule was first discussed by Berry
\cite{Berry:Ozone:1960}.

\begin{figure}[h!]
\includegraphics[width = 8 cm]{./figures/OzonePES3fold.pdf}
\caption{Two-dimensional slice of the potential energy surface of ozone in hyperspherical coordinates at the value of hyperradiuns $\rho$=4 a.u. (the lowest contour is at 1 eV below the dissociation limit). The slice shows three equivalent minima corresponding to 
C$_{2v}$ structures separated by large barriers. 
Reproduced with permission from Ref. \citenum{Babikov:Ozone:2003}.
    \label{fig:3fold}}
\end{figure}

Beyond the structural aspects of ozone, other mysteries surround this
curious molecule. For example, the distribution of the eighteen distinct
$^{16}$O/$^{17}$O/$^{18}$O isotopologues in the Earth's atmosphere differs
from what is expected based on the natural isotopic abundance, a puzzle that
has been open for more than three decades
\cite{Mauersberger:OzoneMystery:1990,Babikov:Ozone:2003}.

Among quantum chemists, ozone has a notorious history, owing to its strong
diradical character causing significant difficulties in calculations of its
ostensibly simple ground-state molecular properties. An early 1989 study
\cite{Stanton:Ozone:1989, Stanton:Ozone:1989b} by the Bartlett group and
collaborators found that the CCSD+T(CCSD) method predicted that the molecular
equilibrium structure of ozone would have C$_s$ symmetry (that is, the
asymmetric stretching harmonic frequency computed by this method is
imaginary), a finding that led to a search for better treatment of
non-iterative triple excitations, ultimately leading to the well-known CCSD(T)
method \cite{Raghavachari:89, Urban:ccsd(t):1985, GenrefCCSD(T):93}. While
CCSD(T) and higher-level coupled-cluster methods available today
\cite{kucharski:ccsdtq:1992, Kallay:CCHigh, Matthews:ncc:2015} do a good job
in describing the equilibrium structure and vibrational potential, an elaborate multireference configuration interaction calculations can describe 
the entire ground-state surface out to the dissociation limit
\cite{Schinke:Ozone:2001,Babikov:Ozone:2003,Dawes:ozone:2013}, facilitating calculations of spectroscopy and chemical reactions of ozone.
As such, the quantum chemical understanding and
fidelity for the ground electronic state of O$_3$ is now at a mature level.

\begin{figure}[h!]
\includegraphics[width = 6cm]{./figures/MOs.pdf}
\caption{Frontier molecular orbitals of ozone; HF/6-31G*. The lowest states of the ozone 
cation are $^2$A$_1$ and $^2$B$_2$ derived by ionization from the HOMO-1 and HOMO-2, respectively.
    \label{fig:MOs}}
\end{figure}

Qualitatively, the challenge posed to electronic structure theory by ozone
ultimately arises from its closely spaced highest-occupied ($a_2$) and
lowest-unoccupied ($b_1$) molecular orbitals (HOMO and LUMO, respectively; see Fig. \ref{fig:MOs}), giving rise to the following wavefunction:
\begin{eqnarray}
X^1A_1 = C_1 [core]^6 (3a_1)^2 (2b_2)^2 (4a_1)^2  (5a_1)^2 (3b_2)^2  (1b_1)^2   (4b_2)^2 (6a_1)^2 (1a_2)^2 (2b_1)^0 + \\
C_2
[core]^6 (3a_1)^2 (2b_2)^2 (4a_1)^2  (5a_1)^2 (3b_2)^2  (1b_1)^2 (4b_2)^2 (6a_1)^2 (1a_2)^0
(2b_1)^2.
\end{eqnarray}
%The weights of these two leading configurations are $C_1\approx$0.952 and $C_2\approx$0.005, according to the EOM-SF-CCSD (EOM-CCSD with spin-flip) calculations\cite{sfccsd}.
%{\bf changing notations, need to check with John! This is confirmed by SF.}
The two electron configurations $[\cdots]a_2^2 b_1^0$ and $[\cdots]a_2^0 b_1^2$ 
%$[\cdots]b_2^2 a_1^0$ and $[\cdots]b_2^0 a_1^2$ 
mix strongly (the weights of these two leading configurations are $C_1\approx$0.952 and $C_2\approx$0.005,
according to the EOM-SF-CCSD (EOM-CCSD with spin-flip) calculations\cite{sfccsd}), 
posing the aforementioned challenges with (especially
single-reference) quantum-chemical methods.   
A second feature consequence of
the symmetry and energetic proximity of these two orbitals is that the ozone
cation (which is isoelectronic to the NO$_2$ radical) has closely lying
$^2$A$_1$ and $^2$B$_2$ electronic states derived by ionization from the HOMO-1 and HOMO-2, respectively (see Fig. \ref{fig:MOs}).
Like the associated states in NO$_2$,
these two states are plagued by orbital symmetry breaking
\cite{Davidson:SymmBreak:76}, a problem that greatly complicates
their quantum-chemical treatment.  One of the many
accomplishments of the Bartlett group has been the integral role played by
them in the development of equation-of-motion coupled-cluster theory
\cite{Stanton:93:EOMCC, Nooijen:EOMEA:95, Bartlett:CC_review:07} (EOM-CC, also
known as linear response coupled-cluster theory \cite{Koch:90:LinResp}). These
methods provide a very efficient and simple way to treat certain classes of electronic structure
that are often termed ``multireference problems" \cite{Krylov:EOMRev:07,Krylov:OSRev}, and
are ideally suited to studying reactive intermediates, radicals,
diradicals, and electronically excited states. From a somewhat wider
viewpoint, the existence of closely spaced electronic states always carries
the potential for (possibly strong) vibronic coupling, a phenomenon that can
play an important role in molecular dynamics and spectroscopy. 
Indeed, one of
the great successes of EOM-CC methods has been in their ability --- if and only
if combined with vibronic coupling models --- to enable high-quality
simulations of complicated electronic spectra.  Such work provides important
insights into the nature of vibronic coupling in molecular systems, as has been
exemplified by various application studies (for example, see
Refs.~\citenum{Stanton:NO3:07} and~\citenum{Koppel:02}). In the case of ozone cation, 
the two lowest electronic states separated by a small gap of $\sim$0.1 eV are coupled  
by the asymmetric stretch of b$_2$ symmetry (Fig. \ref{fig:vib}).

\begin{figure}[h!]
\includegraphics[width = 8 cm]{./figures/vibrations.pdf}
\caption{Three normal modes of the neutral ozone CCSDT/ANO1. Asymmetric stretching vibration of b$_2$ symmetry can couple the two lowest states of the cation ($^2$A$_1$ and $^2$B$_2$). 
\label{fig:vib}}
\end{figure}

The photoelectron spectrum of ozone has been reported\cite{dyke:O3:74} in 1974. Later,  much higher resolution of the positions of vibronic peaks was obtained with 
pulsed-field-ionization zero-kinetic-energy (PFI-ZEKE) spectroscopy \cite{Willitsch:O3ZEKE:2005}. From the theory side, the vibronic photoelectron spectrum of ozone was modeled with a vibronic Hamiltonian parameterized using multi-reference configuration interaction method
by Tarroni and Carter\cite{Willitsch:O3ZEKE:2005}.

@@

%\begin{figure}
%    \includegraphics[width = 16 cm]{./figures/ozone_intro}
%    \caption{ 
%        Two lowest states of the ozone cation and normal modes of the ozone neutral.
%    }
%    \label{fig:ozone_intro}
%\end{figure}

\section{Theoretical models and computational details}

Ozone is a $C_{2v}$ molecule (following Mulliken's
convention~\cite{Mulliken:55:symnot}, the molecule is placed in the $yz$-plane
and the molecular symmetry axis aligns with the $z$-direction) with three
normal modes: symmetric stretch, symmetric bend, and asymmetric stretch (Fig. \ref{fig:vib}).  
The asymmetric stretch is of $b_2$ symmetry. The two lowest electronic states of
the ozone cation, $^2$A$_1$ and $^2$B$_2$, are very close in energy---separated by the gap 
of mere 0.123 eV (or 990 cm$^{-1}$)---and can be coupled by the b$_2$ vibration, giving rise
to significant vibronic effects in the photoelectron spectrum.

We simulate the vibronic states of the ozone cation using the model
Hamiltonian of K{\"o}ppel, Domcke, and Cederbaum---the KDC
Hamiltonian\cite{Cederbaum:LVC:84,KDC:81,Koppel:CIbookCh7:04}. This is a
multi-state and multi-mode Hamiltonian defined in the basis of diabatic
states. For the ozone cation, the basis comprises two quasi-diabatic
states (related to the $^2$A$_1$ and $^2$B$_2$ states at the ground-state minimum), coupled by one mode (asymmetric 
stretch, mode number 3). 
The model also includes two symmetric modes (modes number 1 and 2). 
The matrix elements of the vibronic Hamiltonian are expanded around the equilibrium geometry
of neutral ozone using the respective dimensionless normal coordinates, 
$Q_i$, and the corresponding harmonic frequencies, $\omega_i$.
The overall form  of the vibronic Hamiltonian is thus: 
\begin{subequations}
    \begin{equation}
        H = H _0 \mathbf{1} +
        \begin{pmatrix}
            V^{(1)}  & \lambda _3 Q _3\\
            \lambda _3 Q _3 & V^{(2)}
        \end{pmatrix},
\end{equation}
where $H _0$ contains kinetic energy operator 
and a potential energy term for the coupling mode 
    \begin{equation}
        H _0 = 
        \frac{1}{2} \left(\sum _{i = 1}^3 
        - \omega _i \frac{\partial ^2}{\partial Q _i ^2 }\right)
        + \frac{1}{2}\omega _3 Q _3 ^2,
\end{equation}
and  $V^{(1)}$ and $V^{(1)}$ describe the two diabatic states expanded over $Q_1$ and $Q_2$
\begin{equation}
        V ^{(\alpha)} = 
        E ^{(\alpha)} 
        + \sum _{i \in \{1,2\}} 
            \kappa ^{(\alpha)} _i Q _i 
        + \sum _{i,j \in \{1,2\}} 
            \kappa ^{(\alpha)} _{ij} Q _i Q _j 
        + \sum _{i,j,k \in \{1,2\}} 
            \kappa ^{(\alpha)} _{ijk} Q _i Q _j Q _k 
        + \sum _{i,j,k,l \in \{1,2\}} 
            \kappa ^{(\alpha)} _{ijkl} Q_i Q _j Q _k Q _l.
    \end{equation}
    \label{eq:KDC}
\end{subequations}
Here $E ^{(\alpha)}$ are the vertical ionization energies
calculated at the equilibrium geometry of the neutral, $\kappa$ are the coefficients of expansion of the potential along the fully symmetric coordinates, 
and $\lambda$ is the linear diabatic coupling.
Fig. \ref{fig:KDC} shows the resulting diabatic potential energy surfaces. 
The displacements along the two symmetric normal modes are clearly visible, 
suggesting vibrational progressions along the bend and symmetric stretch.

\begin{figure}[h!]
\includegraphics[width = 8 cm]{./figures/KDC_PES.png}
\caption{Cuts of potential energy surfaces of the two diabatic states of ozone cation 
(corresponding to the $^2$A$_1$ and $^2$B$_2$ states at the equilibrium geometry of the 
neutral) shown as function of the two stretching modes, $Q_1$ and $Q_2$, at $Q_3$=0.
%The displacements along the two symmetric normal modes are clearly visible. 
% Mode labels are correct
%Energy scale is in wavenumbers.
\label{fig:KDC}}
\end{figure}

We computed the parameters of the KDC Hamiltonian using \emph{ab
initio} CC and EOM-CC methods.~\cite{Bartlett:CC_review:07, Krylov:EOMRev:07, Bartlett:Book:09,Christiansen:EOMRev:11, Bartlet:EOMRev:12, Krylov:OSRev} 
We considered methods in which the CC expansion was truncated to singles and doubles (CCSD), singles, doubles and triples (CCSDT), as well as singles, doubles, triples and quadruples (CCSDTQ).\cite{Matthews:ncc:2015}  
We computed the geometry of the neutral ozone and its harmonic
frequencies and normal coordinates with CCSDT/ANO1.~\cite{Almlof:ANO,Almlof:ANO:1988}

The optimized geometry of ozone has a bond length of 1.270 \AA\ and a
bond angle of 116.9$^\circ$. The two symmetric normal modes have frequencies
1,169~cm$^{-1}$ and 724~cm$^{-1}$, and the asymmetric stretch has a
frequency of 1,129~cm$^{-1}$.

We described the states of the cation using 
EOM-CC for ionization energies (known as EOM-IP-CC).~\cite{StantonGauss:EOMIP:99} 
We used the definition of quasi-diabatic states by Ichino, Gauss, and Stanton\cite{Stanton:EOMIPdeg:09}  based on the  EOM-IP-CC method (EOMIP-CC-QD). 
We computed the linear diabatic coupling $\lambda$ and the expansion coefficients $\kappa$ at the equilibrium geometry of the neutral; $\lambda$ was computed with 
EOM-IP-CCSD-QD/ANO1 and  $\kappa$'s were computed using the
EOM-IP-CCSD/ANO1 energies calculated on a grid. The computed value of the linear diabatic coupling constant $\lambda$ is 1,394~cm$^{-1}$. The complete set of KDC parameters is given in the SI.

We computed the vertical ionization energies using a composite method.  The
starting value is the complete basis set (CBS) extrapolation of the
EOM-IP-CCSDT/cc-pCVnZ energies with n = 5, 6.~\cite{Woon:95:CCBS} These values
were augmented with two corrections: the $\Delta$Q correction in the cc-pwCVTZ
basis set and the relativistic correction calculated using
EOM-IP-CCSD/cc-pwCVTZ.~\cite{Dunning:02:p(w)CVnZ} 
The error bars were estimated as follows: 
for the extrapolated CBS reference values, we used half of the absolute value
of the difference between the best \emph{ab initio} value and the extrapolated
value, and for the remaining corrections (quadruples correction, $\Delta Q$, and
the relativistic correction) we used half of the
absolute value of the correction.

To compare the simulated photoelectron spectrum to the experimental one, 
we used the
cross-section ratio A$_1$:B$_2$ of $1$:$1.35$.\cite{KDC:O3:92}
Additionally, the stick spectrum was broadened with the Lorenzian envelopes
normalized to the feature intensities
\begin{equation}
    f(x, x _i, I _i) = 
    \frac{I_i}{(x-x _i)^2 + (\gamma/ 2) ^2},
    \label{eq:lorentzian}
\end{equation}
where $x_i$ is the position of the spectral peak, $I_i$ is its intensity and
$\gamma$ is the peak's width.

The spectrum was computed using the \textsc{xsim} module in \textsc{CFOUR},
with the basis of 50 harmonic states in each mode and 6,000 iterations of the
Lanczos procedure.~\cite{Sharma:xsim_socjt:2024} All electrons were correlated in all CC/EOM-CC calculations. All calculations were performed 
using \textsc{CFOUR}.~\cite{cfour, cfour:2020}

\section{Results and Discussion}

We begin by discussing the accuracy of the computed electronic states.
The vertical ionization energy for the first excited state, $E^{(1)}$, 
computed using the composite
method described above, is  12.827~eV. Our error estimate for this value is
30~meV (see Table~\ref{tab:vertical_ionization_energy} for details).  We note
that the convergence of the vertical gap between the two states is much faster
and this value is estimated as 123$\pm8$~meV (see Table~\ref{tab:vertical_gap}
for details).

\begin{table}[h!]
\caption{
Vertical ionization energies with the error estimates, eV.
    \label{tab:vertical_ionization_energy}}
    \begin{center}  
        \begin{tabular}[c]{|l|rr|r|}
            \hline
            Contribution  & $^2$A$_1$ & $^2$B$_2$ & Error estimate \\ \hline
            EOM-CCSDT/CBS & 12.872    & 12.981    & 0.02 \\
            $\Delta$Q/pwCVTZ & -0.037    & -0.021    & 0.02 \\
            Relativistic/CCSD/pwCVTZ  & -0.008    & -0.010    & 0.005 \\ \hline
            Final value, eV           & 12.827    & 12.950    & 0.03 \\ \hline
        \end{tabular}
    \end{center}
\end{table}
